
---
title: "Health and Economic Impact of Weather Events in the US"
author: "zhenning xu"
date: "May 18, 2019"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
---
---
  title: "cssTest"
    output:
    html_document:
      css: custom.css
      toc: yes
---
---
TOC {
  position: fixed;
  left: 0;
  top: 0;
  width: 200px;
  height: 100%;
  overflow:auto;
}
---

## Synopsis
The basic goal of this assignment is to explore the NOAA Storm Database which contains events from 1950 to Nov 2011 and answer some basic questions about severe weather events. Will try to analyze & present my findings in the following areas:

Across the United States, which types of events are most harmful with respect to population health?
Across the United States, which types of events have the greatest economic consequences?
Introduction
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm Database. This Database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Data
The stormData for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

Storm Data[47Mb]

There is also some documentation of the Database available. Here you will find how some of the variables are constructed/defined.

National Weather Service Storm Data Documentation
National Climatic stormData Center Storm Events FAQ
The events in the Database start in the year 1950 and end in November 2011. In the earlier years of the Database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.


## Data Processing

The analysis was performed on
[Storm Events Database](http://www.ncdc.noaa.gov/stormevents/ftp.jsp), provided by
[National Climatic Data Center](http://www.ncdc.noaa.gov/). The data is from a comma-separated-value file available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).
There is also some documentation of the data available
[here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

The first step is to read the data into a data frame.
```{r cache=TRUE}
setwd("C:/Users/xzhenning/Documents/R/spring 2019/")
storm <- read.csv("C:./data/repdata_data_StormData.csv.bz2")
```

Before the analysis, the data need some preprocessing. Event types don't have a
specific format. For instance, there are events with types `Frost/Freeze`,
`FROST/FREEZE` and `FROST\\FREEZE` which obviously refer to the same type of
event.


```{r}
# number of unique event types
length(unique(storm$EVTYPE))
# translate all letters to lowercase
event_types <- tolower(storm$EVTYPE)
# replace all punct. characters with a space
event_types <- gsub("[[:blank:][:punct:]+]", " ", event_types)
length(unique(event_types))
# update the data frame
storm$EVTYPE <- event_types
```

No further data preprocessing was performed although the event type field can be
processed further to merge event types such as `tstm wind` and `thunderstorm wind`. 

After performing data cleaning, as expected, the number of unique event types reduce
significantly. For further analysis, the cleaned event types are used.


## Dangerous Events with respect to Population Health

To find the event types that are most harmful to population health, the number
of casualties are aggregated by the event type.

```{r}
library(plyr)
casualties <- ddply(storm, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))
# Find events that caused most death and injury
fatal_events <- head(casualties[order(casualties$fatalities, decreasing = T), ], 10)
injury_events <- head(casualties[order(casualties$injuries, decreasing = T), ], 10)
```

Top 10 events that caused largest number of deaths are

```{r}
fatal_events[, c("EVTYPE", "fatalities")]
```

Top 10 events that caused most number of injuries are

```{r}
injury_events[, c("EVTYPE", "injuries")]
```

## Economic Effects of Weather Events

To analyze the impact of weather events on the economy, available property
damage and crop damage reportings/estimates were used.

Note: in the raw data, the property damage is represented with two fields, a number
`PROPDMG` in dollars and the exponent `PROPDMGEXP`. Similarly, the crop damage
is represented using two fields, `CROPDMG` and `CROPDMGEXP`. The first step in the
analysis is to calculate the property and crop damage for each event.

```{r}
exp_transform <- function(e) {
    # h -> hundred, k -> thousand, m -> million, b -> billion
    if (e %in% c('h', 'H'))
        return(2)
    else if (e %in% c('k', 'K'))
        return(3)
    else if (e %in% c('m', 'M'))
        return(6)
    else if (e %in% c('b', 'B'))
        return(9)
    else if (!is.na(as.numeric(e))) # if a digit
        return(as.numeric(e))
    else if (e %in% c('', '-', '?', '+'))
        return(0)
    else {
        stop("Invalid exponent value.")
    }
}
```

```{r cache=TRUE}
prop_dmg_exp <- sapply(storm$PROPDMGEXP, FUN=exp_transform)
storm$prop_dmg <- storm$PROPDMG * (10 ** prop_dmg_exp)
crop_dmg_exp <- sapply(storm$CROPDMGEXP, FUN=exp_transform)
storm$crop_dmg <- storm$CROPDMG * (10 ** crop_dmg_exp)
```


```{r}
# Compute the economic loss by event type
library(plyr)
econ_loss <- ddply(storm, .(EVTYPE), summarize,
                   prop_dmg = sum(prop_dmg),
                   crop_dmg = sum(crop_dmg))
# filter out events that caused no economic loss
econ_loss <- econ_loss[(econ_loss$prop_dmg > 0 | econ_loss$crop_dmg > 0), ]
prop_dmg_events <- head(econ_loss[order(econ_loss$prop_dmg, decreasing = T), ], 10)
crop_dmg_events <- head(econ_loss[order(econ_loss$crop_dmg, decreasing = T), ], 10)
```

Top 10 events that caused most property damage (in dollars) are as follows

```{r}
prop_dmg_events[, c("EVTYPE", "prop_dmg")]
```

Similarly, the events that caused biggest crop damage are

```{r}
crop_dmg_events[, c("EVTYPE", "crop_dmg")]
```

## Plots

The following plots show top dangerous weather event types.

```{r}
library(ggplot2)
library(gridExtra)
# Set the levels in order
p1 <- ggplot(data=fatal_events,
             aes(x=reorder(EVTYPE, fatalities), y=fatalities, fill=fatalities)) +
    geom_bar(stat="identity") +
    coord_flip() +
    ylab("Total number of fatalities") +
    xlab("Event type") +
 ggtitle ("Top 20 events for total fatalities") +
    theme(axis.text.x=element_text(angle=45,hjust=1))
p2 <- ggplot(data=injury_events,
             aes(x=reorder(EVTYPE, injuries), y=injuries, fill=injuries)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    ylab("Total number of injuries") +
    xlab("Event type") +
 ggtitle ("Top 20 events for total injuries") +
    theme(axis.text.x=element_text(angle=45,hjust=1))
p1
p2
```

Tornadoes cause most number of deaths and injuries among all event types. There 
are more than 5,000 deaths and more than 10,000 injuries in the last 60 years
in US, due to tornadoes. 
The other event types that are most dangerous with respect to population health
are excessive heat and flash floods.

Economic impact of weather events
---------------------------------

The following plot shows the most severe weather event types with respect to
economic cost that they have costed since 1950s.


```{r}
library(ggplot2)
library(gridExtra)
# Set the levels in order
p1 <- ggplot(data=prop_dmg_events,
             aes(x=reorder(EVTYPE, prop_dmg), y=log10(prop_dmg), fill=prop_dmg )) +
    geom_bar(stat="identity") +
    coord_flip() +
    xlab("Event type") +
    ylab("Property damage in dollars (log-scale)") +
   ggtitle ("Top 20 events for total property damage") +
    theme(axis.text.x=element_text(angle=45,hjust=1))
p2 <- ggplot(data=crop_dmg_events,
             aes(x=reorder(EVTYPE, crop_dmg), y=crop_dmg, fill=crop_dmg)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    xlab("Event type") +
    ylab("Crop damage in dollars") + 
  ggtitle ("Top 20 events for total crop damage") +
    theme(axis.text.x=element_text(angle=45,hjust=1))
p1
p2
```

## conlcusions
The data shows that flash floods and thunderstorm winds cost the largest
property damages among weather-related natural diseasters. 

The most severe weather event in terms of crop damage is drought. In the last
50 years, drought has caused more than 10 billion dollars in loss. Other
severe crop-damage-causing event types include floods and hails.

Note:Property damages are converted to logarithmic scales due to large range of values.